---
title: "2025 Lin et al. Supplemental Methods"
author: "Benjamin Lin"
editor: "C Ryan Miller"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 6 
    code_folding: hide
layout: page
subtitle: "C Ryan Miller Lab @ UAB"
affiliation: UAB
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: sentence
---

This R Markdown (RMD) file sets up the downstream analysis R environment for the 2025 Lin manuscript. 

# Setup
This code section starts script timing, loads R libraries, sets file directory paths and seeds for analysis, and imports bm_hEGFR orthologous genes between mouse and human (dataframe with matched ensembl gene_id, mouse gene names, and human gene names). 

Script Timing Start
```{r}
ptm <- proc.time()
```

Docker mount directory
```{r}
Data <- paste0("/Data/")
```

Load Libraries
```{r, message = FALSE}
library(tibble)
library(rtracklayer)
library(biomaRt)
library(dplyr)
library(Biobase)
library(tximport)
library(DESeq2)
library(ggplot2)
library(scales)
library(grid)
library(gridExtra)
library(cowplot)
library(colorRamps)
library(RColorBrewer)
library(pheatmap)
library(tidyverse)
library(BiocParallel)
library(AnnotationDbi)
library(org.Hs.eg.db)
library(EnhancedVolcano)
library(gplots)
library(grDevices)
library(rmarkdown)
library(RSQLite)
library(GSVA)
library(msigdbr)
library(UpSetR)
library(ComplexHeatmap)
library(vsn)
library(gprofiler2)
library(ggpubr)
library(PCAtools)
library(DEP)
library(limma)
library(parallel)
library(gtools)
library(cluster)
library(factoextra)
```

Set.Seed
```{r}
# This will set the random number generator
set.seed(888)
```

bm_hEGFR: Human to mouse genes translation matrix
This object will be used for mouse and human gene translation.
```{r}
bm_hEGFR <- readRDS(paste0(Data,"Input/Figures/bm_hEGFR.rds"))
```

## Directories
```{r}
dir.create(paste0(Data,"Output/"))
dir.create(paste0(Data,"Graphs/"))
```

## Generate kinases.anno
```{r}
# 201006_composite_kinases has 16 extra kinases not originally present in 210706_uniprot... so they were added by hand to match
# 210706_uniprot has 11 kinases not present in 201016_composite but they're not important (sperm motility, pseudogenes, etc. so not adding them to kinases or kinases.anno)
#final number of kinases to proceed: 558

kinases <- read.csv("/Data/Input/MIB-MS/201006_composite_kinases.csv", header = TRUE, fileEncoding="UTF-8-BOM")

kinases.anno <- kinases

kinases.uniprot <- read.csv("/Data/Input/MIB-MS/210706_uniprot_mouse_kinases_with_hEGFR.csv", header = TRUE, fileEncoding="UTF-8-BOM")

kinases.uniprot <- subset(kinases.uniprot, kinases.uniprot$Status=="reviewed")

kinases.anno <- left_join(kinases.anno, kinases.uniprot, by = c("Mouse_symbol" = "gene" ))

setdiff(kinases.uniprot$gene, kinases.anno$Mouse_symbol)

# kinases.anno and kinases.uniprot are 11 different, but it's okay

kinases.anno <- subset(kinases.anno, kinases.anno$Status=="reviewed")
 
setdiff(kinases$Mouse_symbol, kinases.anno$Mouse_symbol)
 
isUnique <- function(vector) {
  return(!any(duplicated(vector)))
}

isUnique(kinases$Mouse_symbol)

isUnique(kinases.anno$Mouse_symbol)
```


# Session Info
```{r}
sessionInfo()
```

# Timing End
```{r}
proc.time() - ptm
```

