---
title: "06_MIBs_DYN_sig"
author: "Benjamin Lin"
editor: "C Ryan Miller"
date: '`r format(Sys.time(), "Last modified: %d %b %Y")`'
output:
  html_document:
    toc: yes
    toc_float: yes
    toc_depth: 6 
    code_folding: hide
layout: page
subtitle: "C Ryan Miller Lab @ UAB"
affiliation: UAB
editor_options: 
  chunk_output_type: inline
  markdown: 
    wrap: sentence
---

This R Markdown (RMD) file performs exploratory data analysis (EDA) for the 2025 Lin manuscript.

Script Timing Start
```{r}
ptm <- proc.time()
```

## Variables
```{r}
# adjusted pvalue
qval <- 0.05
pval <- 0.05
lfc <- .5

# Use max of 4 cores if available.
ncore <- ifelse((detectCores()-2)>=4,4,detectCores()-2)

# Bootstrapped values
# we used B = 10 for demo. Recommended value is ~500. Number of bootstrapped values
BS_N <- 10
```

Set.Seed
```{r}
# This will set the random number generator
set.seed(888)
```

# MIBs_TMT_Clustering data
## Directories
```{r}
# output directory
output_dir <- paste0(Data, "Output/MIBs_TMT_Clustering_sig/")
dir.create(paste0(output_dir))

# Graphs
graph_dir <- paste0(Data, "Graphs/MIBs_TMT_Clustering_sig/")
dir.create(paste0(graph_dir))
```

# Import limma res for each cell line
```{r}

cell_lines <- c("CEv3", "G12",  "G8"  , "G5"  , "G1" ,  "E5"  , "E4")

# Import padj
cell_line_limma_res <- list()
for(i in seq_along(cell_lines)){
  cell_line_limma_res[[cell_lines[i]]] <-
    read.csv(file=paste0(Data,"Output/MIB-DYN/DEP_TMT_EDA/",cell_lines[i],"/Limma_res/All_coi_padj.csv"), row.names = 1)
}

# Annotate sig. If at any timepoint padj < 0.05 it is sig
for(i in seq_along(cell_line_limma_res)){
cell_line_limma_res[[i]][,paste0(names(cell_line_limma_res)[i],"_","sig")] <- ifelse(cell_line_limma_res[[i]][,1]<0.05,
                                  "yes",
                                  ifelse(cell_line_limma_res[[i]][,2]<0.05,
                                  "yes",
                                  ifelse(cell_line_limma_res[[i]][,3]<0.05,
                                  "yes","no")))
}

# Extract sig genes
sig_genes <- list()
for (i in seq_along(cell_line_limma_res)){
  sig_genes[[names(cell_line_limma_res)[i]]] <-
    c(row.names(cell_line_limma_res[[i]][cell_line_limma_res[[i]][,4]=="yes",]))
}


```

## Import TMT counts
```{r}
DEP_TMT_EDA_mv_ls <- readRDS(tail(grep("DEP_TMT_EDA_mv_ls.rds",list.files(paste0(Data,"Output/MIB-DYN/DEP_TMT_EDA"), recursive = T, full.names = T), value = T),1))
DEP_TMT_EDA_mv_ls <- subset(DEP_TMT_EDA_mv_ls, names(DEP_TMT_EDA_mv_ls) %in% cell_lines)
```


# Convert count to Z score
```{r}
# Extract clustering_normCounts and convert to z score
clustering_normCounts_ls <- list()
cluster_mat_ls <- list() # List of matrix
cluster_mat.rn_ls <- list() # Gene names
for(h in seq_along(DEP_TMT_EDA_mv_ls)){
  normCounts <- assay(DEP_TMT_EDA_mv_ls[[h]])
  
  # Convert normCounts to Z score
cluster_mat <- t(scale(t(normCounts)))

# Remove rows with NA
cluster_mat <- na.omit(cluster_mat)

# Extract all genes 
cluster_mat.rn <-row.names(cluster_mat)

  # Save results to list
  clustering_normCounts_ls[[names(DEP_TMT_EDA_mv_ls)[h]]] <- normCounts
  cluster_mat_ls[[names(DEP_TMT_EDA_mv_ls)[h]]] <- cluster_mat
  cluster_mat.rn_ls [[names(DEP_TMT_EDA_mv_ls)[h]]] <- cluster_mat.rn
}

# Extract all genes to list of df with metadata
DEP_TMT_EDA_cluster_mat_gene_ls <-list ()
for (h in seq_along(cluster_mat_ls)){
  # Extract
  cluster_mat <- cluster_mat_ls[[h]]
  cluster_mat.rn <- cluster_mat.rn_ls[[names(cluster_mat_ls)[h]]]
  DEP_TMT_EDA <- DEP_TMT_EDA_mv_ls[[names(cluster_mat_ls)[h]]]
  
# For each se object, extract all genes
cluster_mat_gene_ls <- list()
system.time(for (i in seq_along(cluster_mat.rn)) { 
  cluster_mat_gene_ls[[row.names(cluster_mat)[i]]] <- as.data.frame(cluster_mat[i,])
  colnames(cluster_mat_gene_ls[[cluster_mat.rn[i]]][1]) <- cluster_mat.rn[i]
  cluster_mat_gene_ls[[cluster_mat.rn[i]]] <-
    rownames_to_column(cluster_mat_gene_ls[[cluster_mat.rn[i]]], var = "sample_id")
  colnames(cluster_mat_gene_ls[[cluster_mat.rn[i]]])[2] <- "z_norm_counts"
    cluster_mat_gene_ls[[cluster_mat.rn[i]]]$Cell.Line <-
    DEP_TMT_EDA@colData$Cell.Line[match(cluster_mat_gene_ls[[cluster_mat.rn[i]]]$sample_id,
                                          rownames(DEP_TMT_EDA@colData))]
  cluster_mat_gene_ls[[cluster_mat.rn[i]]]$Timepoint <-
    DEP_TMT_EDA@colData$Timepoint[match(cluster_mat_gene_ls[[cluster_mat.rn[i]]]$sample_id,
                                       rownames(DEP_TMT_EDA@colData))]
  cluster_mat_gene_ls[[cluster_mat.rn[i]]]$cell_line_abbr <-
    DEP_TMT_EDA@colData$cell_line_abbr[match(cluster_mat_gene_ls[[cluster_mat.rn[i]]]$sample_id,
                                  rownames(DEP_TMT_EDA@colData))]
  cluster_mat_gene_ls[[cluster_mat.rn[i]]]$replicate <-
    DEP_TMT_EDA@colData$replicate[match(cluster_mat_gene_ls[[cluster_mat.rn[i]]]$sample_id,
                                  rownames(DEP_TMT_EDA@colData))]
}
)
DEP_TMT_EDA_cluster_mat_gene_ls[[names(cluster_mat_ls)[h]]] <- cluster_mat_gene_ls

# Save as RDS
saveRDS(DEP_TMT_EDA_cluster_mat_gene_ls, file = paste0(output_dir,"DEP_TMT_EDA_cluster_mat_gene_ls",".rds"))

}
```


## Extract Median and SumStats
```{r}
# Function to get medians 
medians_cluster_data <- function(data) {
  data %>%
    group_by(Cell.Line, Timepoint) %>%
    summarise(
      median = median(z_norm_counts, na.rm = TRUE),
      .groups = 'drop'
    )
}

# Function to get stats
stats_cluster_data <- function(data) {
  data %>%
    group_by(Cell.Line, Timepoint) %>%
    summarise(
      n = n(),
      avg_normCount = mean(z_norm_counts),
      median = median(z_norm_counts, na.rm = TRUE),
      sd = sd(z_norm_counts),
      IQR = IQR(z_norm_counts),
      MAD = mad(z_norm_counts),
      Coeff.Variation.Prcnt =sd(z_norm_counts) / mean(z_norm_counts) * 100,
      .groups = 'drop'
    )
}

# Extract median values from Z scores
DEP_TMT_EDA_cluster_mat_gene_ls_stats <- list()
DEP_TMT_EDA_cluster_mat_gene_ls_median <- list()
DEP_TMT_EDA_cluster_mat_gene_median <- list()
z_mat_ls <- list()
DEP_TMT_EDA_cluster_mat_gene_sd <- list()
system.time(for(h in seq_along(DEP_TMT_EDA_cluster_mat_gene_ls)){
  
  # Extract one
  cluster_mat_gene_ls <- DEP_TMT_EDA_cluster_mat_gene_ls[[h]]
  
  # Directories
  dir.create(paste0(output_dir,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"/"))
  output_dir_med <- paste0(output_dir,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"/")
  dir.create(paste0(graph_dir,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"/"))
  graph_dir_med <- paste0(graph_dir,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"/")
  
  # Compute stats and calculate medians
cluster_mat_gene_ls_stats <- list()
cluster_mat_gene_ls_median <- list()

# Apply the median function to each gene
system.time(cluster_mat_gene_ls_median <- lapply(cluster_mat_gene_ls, medians_cluster_data))

# Apply the stats function to each gene
system.time(cluster_mat_gene_ls_stats <- lapply(cluster_mat_gene_ls, stats_cluster_data))

# Save intermediates
  DEP_TMT_EDA_cluster_mat_gene_ls_stats[[names(DEP_TMT_EDA_cluster_mat_gene_ls)[h]]] <- cluster_mat_gene_ls_stats
  DEP_TMT_EDA_cluster_mat_gene_ls_median[[names(DEP_TMT_EDA_cluster_mat_gene_ls)[h]]] <- cluster_mat_gene_ls_median

# Save as RDS
  saveRDS(cluster_mat_gene_ls_stats, file = paste0(output_dir_med,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],".rds"))


# Clean up each gene median DF for full join
for ( i in seq_along(cluster_mat_gene_ls_median)){
  # Rename median column to gene name
  names(cluster_mat_gene_ls_median[[names(cluster_mat_gene_ls)[i]]])[names(cluster_mat_gene_ls_median[[names(cluster_mat_gene_ls)[i]]])=="median"] <- paste0(names(cluster_mat_gene_ls)[i])
  # Metadata column
  cluster_mat_gene_ls_median[[i]]$med_sample <- paste0(cluster_mat_gene_ls_median[[i]]$Timepoint,".",cluster_mat_gene_ls_median[[i]]$Cell.Line)
  # Keep median column and metadata column
  cluster_mat_gene_ls_median[[i]] <- cluster_mat_gene_ls_median[[i]][,which(colnames(cluster_mat_gene_ls_median[[i]]) %in% c(paste0(names(cluster_mat_gene_ls_median)[i]),"med_sample"))]
}

# Convert list to dataframe by timepoints
system.time(cluster_mat_gene_median <- cluster_mat_gene_ls_median %>% reduce(full_join, by = 'med_sample') )

# Oranize and label dataframe
cluster_mat_gene_median <- data.frame(cluster_mat_gene_median)
rownames(cluster_mat_gene_median) <- cluster_mat_gene_median$med_sample

# Remove sample names from matrix
cluster_mat_gene_median <- cluster_mat_gene_median[,!colnames(cluster_mat_gene_median)=="med_sample"]

# Pivot 
t_cluster_mat_gene_median <- t(cluster_mat_gene_median)

# Median values 
DEP_TMT_EDA_cluster_mat_gene_median[[names(DEP_TMT_EDA_cluster_mat_gene_ls)[h]]] <- data.frame(t_cluster_mat_gene_median)

# Save Median Values
write.csv(data.frame(t_cluster_mat_gene_median), file = paste0(output_dir_med,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"_med",".csv"))


# Scale matrix by Z score.
z_mat <- (t_cluster_mat_gene_median-mean(t_cluster_mat_gene_median))/sd(t_cluster_mat_gene_median)

# Save z Values
write.csv(z_mat, file = paste0(output_dir_med,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"_z_mat",".csv"))

# Z matrix for clustering
z_mat_ls[[names(DEP_TMT_EDA_cluster_mat_gene_ls)[h]]] <- z_mat

# Clean up sd stat for full join
for ( i in seq_along(cluster_mat_gene_ls_stats)){
  # Rename sd column to gene name
  names(cluster_mat_gene_ls_stats[[names(cluster_mat_gene_ls)[i]]])[names(cluster_mat_gene_ls_stats[[names(cluster_mat_gene_ls)[i]]])=="sd"] <- paste0(names(cluster_mat_gene_ls)[i])
  # Metadata column
  cluster_mat_gene_ls_stats[[i]]$sd_sample <- paste0(cluster_mat_gene_ls_stats[[i]]$Timepoint,".",cluster_mat_gene_ls_stats[[i]]$Cell.Line)
  # Keep median column and metadata column
  cluster_mat_gene_ls_stats[[i]] <- cluster_mat_gene_ls_stats[[i]][,which(colnames(cluster_mat_gene_ls_stats[[i]]) %in% c(paste0(names(cluster_mat_gene_ls_stats)[i]),"sd_sample"))]
}

# Convert list to dataframe by timepoints
system.time(cluster_mat_gene_stats <- cluster_mat_gene_ls_stats %>% reduce(full_join, by = 'sd_sample') )

# Organize and label dataframe
cluster_mat_gene_stats <- data.frame(cluster_mat_gene_stats)
rownames(cluster_mat_gene_stats) <- cluster_mat_gene_stats$sd_sample

# Remove sample names from matrix
cluster_mat_gene_stats <- cluster_mat_gene_stats[,!colnames(cluster_mat_gene_stats)=="sd_sample"]

# Pivot 
t_cluster_mat_gene_stats <- data.frame(t(cluster_mat_gene_stats))

# Save SD
DEP_TMT_EDA_cluster_mat_gene_sd[[names(DEP_TMT_EDA_cluster_mat_gene_ls)[h]]] <- t_cluster_mat_gene_stats

# Save SD Values
write.csv(t_cluster_mat_gene_stats, file = paste0(output_dir_med,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"_sd",".csv"))

})


```


## Filter Z matrix by sig genes
```{r}


# Filter z matrix
z_mat_filt_ls <- list()
for (h in seq_along(z_mat_ls)){
  
  # Directories
  dir.create(paste0(output_dir,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"/"))
  output_dir_z <- paste0(output_dir,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"/")
  dir.create(paste0(graph_dir,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"/"))
  graph_dir_z <- paste0(graph_dir,names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],"/")
  
  
  # Extract one
  z_mat <- z_mat_ls[[h]]
  t_cluster_mat_gene_stats <- DEP_TMT_EDA_cluster_mat_gene_sd[[names(z_mat_ls)[h]]]
  
  # Save Z mat
  write.csv(z_mat, file = paste0(output_dir_z,"z_",names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],".csv"))

  
  # Filter
  z_mat_filt <- z_mat[row.names(z_mat) %in% (sig_genes[[paste0(names(DEP_TMT_EDA_cluster_mat_gene_ls)[h])]]),]

  # Save
  z_mat_filt_ls[[names(z_mat_ls)[h]]] <- data.frame(z_mat_filt)
  
  # Save Z filt
  write.csv(data.frame(z_mat_filt), file = paste0(output_dir_z,"z_filt_",names(DEP_TMT_EDA_cluster_mat_gene_ls)[h],".csv"))

}

# Save as RDS
saveRDS(z_mat_ls, file = paste0(output_dir,"z_mat_ls",".rds"))
saveRDS(z_mat_filt_ls, file = paste0(output_dir,"z_mat_filt_ls",".rds"))


```

#### K-means clustering
```{r}
# K-means clustering (2 clusters)
kmeans_res_ls <- list()
for (i in seq_along(z_mat_filt_ls)){
  set.seed(888)
  kmeans_res_ls[[names(z_mat_filt_ls)[i]]] <-
    kmeans(z_mat_filt_ls[[i]],2, nstart = 25)
}


# Merge clusters with gene matrix
z_mat_clut_clid_ls <- list()
for (i in seq_along(z_mat_filt_ls)){
  
  # Directories
  dir.create(paste0(output_dir,names(z_mat_filt_ls)[i],"/"))
  output_dir_clut <- paste0(output_dir,names(z_mat_filt_ls)[i],"/")
  dir.create(paste0(graph_dir,names(z_mat_filt_ls)[i],"/"))
  graph_dir_clut <- paste0(graph_dir,names(z_mat_filt_ls)[i],"/")
  
  z_mat_clut_clid_ls[[names(z_mat_filt_ls)[i]]] <-
    data.frame(cbind(z_mat_filt_ls[[names(z_mat_filt_ls)[i]]], cluster = as.factor(kmeans_res_ls[[names(z_mat_filt_ls)[i]]]$cluster)))

# Save DF
write.csv(z_mat_clut_clid_ls[[names(z_mat_filt_ls)[i]]], file = paste0(output_dir_clut,"ClusterGenes_",names(z_mat_clut_clid_ls)[i],".csv"))
}


# Assign up or down label
for (i in seq_along(z_mat_clut_clid_ls)){

z_mat_clut_clid <- z_mat_clut_clid_ls[[i]]

z_mat_clut_clid <- data.frame(z_mat_clut_clid)
z_mat_clut_clid$gene <- row.names(z_mat_clut_clid)
# Get mean z score of 0 and 48
cluster_assign <- z_mat_clut_clid %>% group_by(cluster) %>% summarise(mean_0=mean(.data[[grep("X0",colnames(z_mat_clut_clid),value = T)]]),
                                                            mean_48=mean(.data[[grep("X48",colnames(z_mat_clut_clid),value = T)]]))
# Get difference (48-0)
cluster_assign$difference <- (cluster_assign$mean_48-cluster_assign$mean_0)

# Assign up and down labels
cluster_assign$label <- ifelse(cluster_assign$difference>0,"up","down")

# Add labels
z_mat_clut_clid <- inner_join(z_mat_clut_clid,data.frame(cluster_assign), by="cluster")

# Add back rownames
row.names(z_mat_clut_clid) <- z_mat_clut_clid$gene

# Overwrite
z_mat_clut_clid_ls[[names(z_mat_clut_clid_ls)[i]]] <- z_mat_clut_clid
}

# Save RDS of labeled genes and clusters
saveRDS(z_mat_clut_clid_ls, file = paste0(output_dir,format(Sys.Date(),"%y%m%d"),"_","z_mat_clut_clid_ls",".rds"))

```

# Heatmap of clusters
```{r}
ggplot_df_ls <- list()
for (h in seq_along(DEP_TMT_EDA_mv_ls)){
 
kmeans_res_clusters <- z_mat_clut_clid_ls[[names(DEP_TMT_EDA_mv_ls)[h]]]
  
# Values are already transformed. Extract to vsd
vsd <- DEP_TMT_EDA_mv_ls[[h]]
head(assay(vsd),3)

# Subset dds to contain genes in the clustering analysis
cluster_vsd <-vsd[row.names(vsd) %in% row.names(kmeans_res_clusters),]

# Heatmap_GOI_df_ls
heatmap_goi <- kmeans_res_clusters

# Extract matrix values
mat_list <- as.matrix(assay(cluster_vsd))

# Heatmap annotation labels
annotation_col <- HeatmapAnnotation(df = data.frame(Model = cluster_vsd@colData$Timepoint),
    col = list(time_h= 
                 c(
    '0'='black',
    '4'='blue',
    '24'='purple',
    '48'='orange'
  )
  ))

# Generate Heatmap
Heatmap <- ComplexHeatmap::Heatmap(
  t(scale(t(mat_list))),
  top_annotation = annotation_col,
  name = " ",
  column_split = cluster_vsd@colData$Timepoint,
  column_order = colnames(cluster_vsd)[order(cluster_vsd@colData$Timepoint)],
  column_title = paste0(names(DEP_TMT_EDA_mv_ls)[h]," MIBs 3 µM Afatinib"),
  row_split = heatmap_goi[row.names(mat_list),]["label"] ,
  row_gap = unit(1, "mm"),
  show_column_names = F,
  show_row_names = T,
  row_names_gp = gpar(fontsize = 8),
  width = ncol(mat_list)*unit(5, "mm"), 
  height = nrow(mat_list)*unit(2.5, "mm")
  )

draw(Heatmap)

# Directories
  dir.create(paste0(output_dir,names(DEP_TMT_EDA_mv_ls)[h],"/"))
  output_dir_heatmap <- paste0(output_dir,names(DEP_TMT_EDA_mv_ls)[h],"/")
  dir.create(paste0(graph_dir,names(DEP_TMT_EDA_mv_ls)[h],"/"))
  graph_dir_heatmap <- paste0(graph_dir,names(DEP_TMT_EDA_mv_ls)[h],"/")

# Get height and width of heatmap for pdf
width = convertX(ComplexHeatmap:::width(draw(Heatmap)), "inch", valueOnly = TRUE)
height = convertY(ComplexHeatmap:::height(draw(Heatmap)), "inch", valueOnly = TRUE)

pdf(file=paste0(graph_dir_heatmap, "ClusterHeatmap_" , names(DEP_TMT_EDA_mv_ls)[h], ".pdf"), width = width, height = height)
draw(Heatmap)
dev.off()
  
}

```

## Gene Profile Plots
```{r}
for (h in seq_along(z_mat_clut_clid_ls)){
  

# Extract one
kmeans_res_clusters <- z_mat_clut_clid_ls[[names(z_mat_clut_clid_ls)[h]]]
z_mat_filt <- z_mat_filt_ls[[names(z_mat_clut_clid_ls)[h]]]


# Organization for ggplot dataframe
df <- data.frame(z_mat_filt)
df$gene <- row.names(df)
df_ggplot <- df %>% pivot_longer(-gene)
cluster <- data.frame(kmeans_res_clusters[,"label"])
cluster$gene <- row.names(kmeans_res_clusters)
row.names(cluster) <- cluster$gene
df_cluster_ggplot <-  inner_join(df_ggplot, cluster, by = "gene")
names(df_cluster_ggplot)[4] <- "label"
df_cluster_ggplot$Timepoint <- gsub("X","",sapply(strsplit(df_cluster_ggplot$name,"\\."),"[",1))

gene_plot <- df_cluster_ggplot %>%
ggplot(aes(Timepoint, value)) +
  geom_line(aes(group = gene)) +
  facet_wrap(vars(label), ncol = 4) + 
    labs(title = paste0("MIB Clusters ",names(z_mat_clut_clid_ls)[h])) + 
  theme_bw() +
  scale_x_discrete(limits = c("0", "4", "24", "48"), expand = c(0,0.1))

plot(gene_plot)

# Save plot

# Directories
  dir.create(paste0(output_dir,names(z_mat_clut_clid_ls)[h],"/"))
  output_dir_gp <- paste0(output_dir,names(z_mat_clut_clid_ls)[h],"/")
  dir.create(paste0(graph_dir,names(z_mat_clut_clid_ls)[h],"/"))
  graph_dir_gp <- paste0(graph_dir,names(z_mat_clut_clid_ls)[h],"/")

pdf(file=paste0(graph_dir_gp, "GeneProfilePlot_" , names(z_mat_clut_clid_ls)[h], ".pdf"))
plot(gene_plot)
dev.off()

}
```

# Session Info
```{r}
sessionInfo()
```

# Timing End
```{r}
proc.time() - ptm
```


